---
title: "The Zoom Boom: How Remote Work Rewired Where Americans Live"
author: "Hue Khoi Ly"
format: revealjs
theme: solarized
slide-number: true
editor: visual
execute: 
  echo: false
  warning: false
  message: false
---

```{r}
# Load packages needed for the slides.
#| include: false
library(tidyverse)
library(jsonlite)
library(tidycensus)
library(ggplot2)
library(maps)
library(ggrepel)
```

## Research Question

::: incremental
How did the rise of remote work after 2020 change where Americans live?

This project contains three parts of the story:

1.  Remote work stayed above pre-pandemic levels.
2.  Population and domestic migration shifted towards several Sun Belt and Mountain West states.
3.  These migration patterns may have increased housing pressure in destination communities.
:::

## Part 1: Remote Work {.smaller}

::: {style="font-size: 0.85em;"}
Following the COVID-19 pandemic, the rise of remote work fundamentally transformed how people operate and communicate professionally. With the rapid development of remote-work platforms and applications, employees can easily connect, share data, and collaborate without needing to be physically present at the office. This shift raises an important question: Has this trend been a root cause of migration to more affordable states, subsequently driving up living costs and making those regions unaffordable? This research aims to investigate and establish the relationship between these two phenomena. To begin, we will analyze whether current remote work rates remain significantly higher than pre-pandemic levels.

Data source: U.S. Census ACS, WFH Research – U.S. Survey of Working Arrangements and Attitudes (SWAA)
:::

## 1. How much did remote work change after COVID-19? {.smaller}

```{r}

#pull data from the JSON link
x <- fromJSON("https://wfhresearch.com/datafiles/WFHtimeseries_monthly.json", simplifyVector = FALSE)

#choose main series in the list
main_series <- x$series[[which(
  sapply(x$series, \(s) s$source_id) == "wfhcovid_matquestion"
)]]

#change the list to data frame
remote_tbl <- as.data.frame(do.call(rbind, main_series$observations))
names(remote_tbl) <- c("date", "value")

#change the type of the column value
rm_USA <- remote_tbl |>
  mutate(
    date = as.Date(unlist(date)),
    value = as.numeric(unlist(value))
  )

baseline <- rm_USA |>
  filter(date == as.Date("2019-12-01")) |>
  pull(value)
```

:::::: columns
::: {.column width="70%"}
```{r}
#create line graph for first question
rm_USA |> 
  ggplot(aes(date,value))+
  geom_line()+
  geom_hline(yintercept = baseline, linetype = "dashed", color = "red") +
  labs(
    title = "SWAA work from home rate",
    x = "Date",
    y = "Percent of full paid working days"
  )
```
:::

:::: {.column width="30%"}
::: {style="font-size: 0.65em;"}
The remote work surged dramatically during the COVID shock, then fell from its emergency peak. However, it stabilized at a level (26%) still far above the pre-pandemic baseline (December 1st 2019) `{r} baseline`

*Reference link: Barrero, Jose Maria, Nicholas Bloom, and Steven J. Davis, 2021. “Why working from home will stick,” National Bureau of Economic Research Working Paper 28731.*
:::
::::
::::::

## 2. Which Western and Sun Belt states had the highest remote-work shares? {.smaller}

```{r}
#| include: false
# pull data from census
#census_api_key("API key",install = "TRUE") before get_acs()
rm.pc_24<- get_acs(
  geography = "state",
  variables = "DP03_0024PE",
  survey = "acs1",
  year = 2024,
  geometry = TRUE
)
target_states <- c("California", "Nevada", "Arizona", "Texas", "Idaho","Colorado", "Utah", "Florida")
rm.pc_24.tg <- rm.pc_24 |> 
  filter(NAME %in% target_states)
```

```{r}
# create bar chart for question 2:
rm.pc_24.tg |> ggplot(aes(x=reorder(NAME,estimate),y= estimate))+
  geom_col(fill = '#00BF63')+
  coord_flip() +
  labs(
    title = "Share of workers working from home",
    x = "",
    y = "Percent"
  )
```

::: {style="font-size: 0.65em;"}
Among the selected Western and Sun Belt states, Colorado had the highest work-from-home share, followed by Arizona, while Nevada had the lowest share. California is in the middle of the list. But I think with the total amount of jobs in California, that percentage represent that there are a lot of remote-work in California.
:::

## 3. Which states had the biggest growth in remote work? {.smaller}

```{r}
#|include: false
##pull data
years <- c(2019, 2021, 2022, 2023, 2024)
remote_by_year <- lapply(years, function(y) {
  get_acs(
    geography = "state",
    variables = "DP03_0024PE",
    year = y,
    survey = "acs1"
  ) |>
    mutate(year = y)
}) |>
  bind_rows()

## wider the table and filter the table
remote_growth <- remote_by_year |>
  filter(NAME %in% target_states) |>
  select(NAME, year, estimate) |>
  pivot_wider(names_from = year, values_from = estimate) |>
  mutate(change_2019_2024 = `2024` - `2019`)
```

```{r}
## create lolipop graph for q3:
ggplot(remote_growth, aes(x = reorder(NAME, change_2019_2024), y = change_2019_2024)) +
  geom_segment(aes(xend = NAME, y = 0, yend = change_2019_2024), color = "gray60") +
  geom_point(size = 3, color = "steelblue") +
  coord_flip() +
  labs(
    title = "Growth in remote-work share, 2019 to 2024",
    x = "",
    y = "Percentage-point change"
  ) +
  theme_minimal()
```

::: {style="font-size: 0.65em;"}
Remote-work growth between 2019 and 2024 was strongest in Colorado and Arizona, while Idaho saw the smallest.
:::

## 4. In California, how did remote work compare with onsite work from 2019 to now? {.smaller}

```{r}
#| include: false
#pull data 
ca_remote <- lapply(years, function(y) {
  get_acs(
    geography = "state",
    variables = "DP03_0024PE",
    year = y,
    survey = "acs1"
  ) |>
    filter(NAME == "California") |>
    mutate(year = y)
}) |>
  bind_rows()
#create another column that is onsite
ca_compare <- ca_remote |>
  transmute(
    year,
    remote = estimate,
    onsite = 100 - estimate
  )
# longer the data table
ca_long <- ca_compare |>
  pivot_longer(
    cols = c(remote, onsite),
    names_to = "work_type",
    values_to = "percent"
  )
```

```{r}
# draw the stacked bar charts 
ggplot(ca_long, aes(x = factor(year), y = percent, fill = work_type)) +
  geom_col() +
  labs(
    title = "Remote vs onsite work in California",
    x = "Year",
    y = "Percent of workers",
    fill = "Work type"
  ) +
  theme_minimal()
```

::: {style="font-size: 0.65em;"}
In California, remote work rose sharply after 2019, peaked in 2021 due to the covid-19 pandemic. Onsite work partially recovered but remained below its pre-pandemic share.
:::

## 5. Did remote work grow more in high-cost states than in lower-cost destination states? {.smaller}

```{r}
#| include: false
#longer the table
remote_long <- remote_growth |>
  select(NAME, `2019`, `2021`, `2022`, `2023`, `2024`) |>
  pivot_longer(
    cols = c(`2019`, `2021`, `2022`, `2023`, `2024`),
    names_to = "year",
    values_to = "remote_share"
  ) |>
  mutate(year = as.numeric(year))
```

```{r}
# create line plot
ggplot(remote_long, aes(x = year, y = remote_share, color = NAME, group = NAME)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Remote-work share by year",
    x = "Year",
    y = "Percent working from home",
    color = "State"
  ) +
  theme_minimal()
```

::: {style="font-size: 0.65em;"}
The current results suggest that remote-work growth was not concentrated only in high-cost states, since several lower-cost destination states also showed large increases.
:::

## 6. Did places with more remote work also gain population? {.smaller}

```{r}
#| include: false
# pull population data variable from census
pop_2019 <- get_acs(
  geography = "state",
  variables = "B01003_001",
  year = 2019,
  survey = "acs1"
) |>
  select(GEOID, NAME, pop_2019 = estimate)

pop_2024 <- get_acs(
  geography = "state",
  variables = "B01003_001",
  year = 2024,
  survey = "acs1"
) |>
  select(GEOID,NAME , pop_2024 = estimate)
# create the table with the population change column
pop_change <- pop_2019 |>
  left_join(pop_2024, by = c("GEOID", "NAME")) |>
  mutate(
    pop_growth_pct = (pop_2024 - pop_2019) / pop_2019 * 100
  )
#combine together
state_compare <- remote_growth |>
  left_join(pop_change, by = "NAME")
```

```{r}
#draw the scatter plot with the line
ggplot(state_compare, aes(x = change_2019_2024, y = pop_growth_pct)) +
  geom_point(size = 3, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick") +
  geom_text_repel(
    data = subset(state_compare, NAME %in% target_states),
    aes(label = NAME),
    size = 4
  ) +
  labs(
    title = "Remote-work growth and population growth by state",
    x = "Change in remote-work share, 2019 to 2024",
    y = "Population growth (%), 2019 to 2024"
  ) +
  theme_minimal()
```

::: {style="font-size: 0.65em;"}
The remote-work growth does not appear to be positively associated with population growth; the fitted line is negative, and states such as Idaho and California do not fit a simple “more remote work means more population gain” pattern. However, I think people can get a remote job from one state that has a high remote-work growth rate and move to another state that is affordable.
:::

::: ::::::
:::
